---
title: "copy_number_analysis_TCGA_ACC"
author: "Evan_Naughton"
date: "2024-05-13"
output: html_document
---

```{r}
library(tidyr)
library(tidyverse)
library(ggplot2)
library(dplyr)
```

Values: -2 = homozygous deletion; -1 = hemizygous deletion; 0 = neutral / no change; 1 = gain; 2 = high level amplification.

  -2 or Deep Deletion indicates a deep loss, possibly a homozygous deletion
  -1 or Shallow Deletion indicates a shallow loss, possibley a heterozygous deletion
   0 is diploid
   1 or Gain indicates a low-level gain (a few additional copies, often broad)
   2 or Amplification indicate a high-level amplification (more copies, often focal)

```{r}
# Reading in the clinical data

finaldf <- read.table("/home/evannaughton/project/r_code/finaldf.txt", header = T)
head(finaldf)
```

```{r}
# Reading in the copy number data

copy_number_data <- read.delim("/home/evannaughton/project/acc_tcga/data_cna.txt", header = T)
head(copy_number_data)
```

```{r}
# reading in the other copy number data

copy_number_data_hg19 <- read.delim("/home/evannaughton/project/acc_tcga/data_cna_hg19.seg", header = T)
head(copy_number_data_hg19)
```

```{r}
# Cycling through copy number data to obtain MKI67 copy number for each patient

# Identifying the MKI67 gene
mki67 <- which(copy_number_data$Hugo_Symbol == "MKI67")

# Extracting MKI67 copy number values for each sample
mki67_cn <- copy_number_data[mki67, ]  

mki67_cn <- mki67_cn[, -2]

print(mki67_cn)
```

```{r}
# Pivoting to long format

long_df <- pivot_longer(mki67_cn, 
                         cols = -Hugo_Symbol, 
                         names_to = "Sample.ID",  
                         values_to = "MKI67.Copy.Number")  

long_df <- long_df[, c(2, 1, 3)]

head(long_df)
```

```{r}
# Merging dataframes

# converting sample IDs to suitable format
long_df$Sample.ID <- gsub("\\.", "-", long_df$Sample.ID)

# removing Hugo_Symbol column
long_df <- subset(long_df, select = -Hugo_Symbol)

# Merged dataframes based on Sample_ID
finaldf <- merge(finaldf, long_df, by = "Sample.ID", all = TRUE)

# Remove rows with NA only in the Study.ID column
finaldf <- subset(finaldf, !is.na(Study.ID))

# merging dataframes based on MKI67.Copy.Number resulted in the loss of two samples as their MKI67 copy number information was not available (i.e. NA). Overcame this by merging all 91 samples with the finaldf and then removing the samples with NA in the Study.ID column. This resulted in a dataframe with all 79 patients with their respective MKI67 copy number along with the 2 samples with missing information for this variable.

# Reset row indices
rownames(finaldf) <- NULL
```

```{r}
# reading in segmented copy number data file

segmented_copy_number_data <- read.delim("/home/evannaughton/project/acc_tcga/data_cna_hg19.seg", header = T)
head(segmented_copy_number_data)
```

```{r}
# Separating the data based on quartiles

quartile_data <- finaldf %>% 
  filter(quartile %in% c(1, 4))

# Turning quartile into a factor

quartile_data$quartile <- as.factor(quartile_data$quartile)

# Turning MKI67 CNA into a factor

quartile_data$MKI67.Copy.Number <- as.factor(quartile_data$MKI67.Copy.Number)
```

```{r}
# Bar plot comparing MKI67 CNA score on the x-axis with the categorical variable "Quartile" on the y-axis
p <- ggplot(quartile_data, aes(x = factor(MKI67.Copy.Number), fill = factor(quartile))) +
  geom_bar(position = position_dodge(preserve = "single"), color = "black", width = 0.7) +
  scale_fill_manual(values = c("turquoise", "red"), name = "quartile") +
  ggtitle("MKI67 CNA score vs. Sample Count For Each Quartile (1 = low; 4 = high)") +
  labs(x = "MKI67 CNA Score", y = "Count For Each Quartile", 
       caption = "-2 = homozygous deletion; -1 = hemizygous deletion; 0 = neutral / no change; 1 = gain; 2 = high level amplification") +
  theme_minimal() +
  theme(legend.position = "top")  # Adjust legend position

# Display the plot
print(p)
```

The fourth quartile (high MKI67 expression) is associated with higher genomic instability within the MKI67 gene than the low MKI67 expression quartile.

PERFORMING DIFFERENTIAL COPY NUMBER ANALYSIS BETWEEN HIGH AND LOW GROUPS
```{r}
# Separating the data based on quartiles

low_samples <- finaldf %>% 
  filter(quartile %in% c(1))

# Separating the data based on quartiles

high_samples <- finaldf %>% 
  filter(quartile %in% c(4))

# Changing sample IDs to match the IDs in gdata

low_samples$Sample.ID <- gsub("-", "\\.", low_samples$Sample.ID)
head(low_samples)

high_samples$Sample.ID <- gsub("-", "\\.", high_samples$Sample.ID)
head(high_samples)
```

```{r}
# Subset the data into high and low groups based on MKI67 expression levels

# Identify Sample IDs in data_cna
sample_ids_cna <- colnames(copy_number_data)[-c(1, 2)]  

# Match sample IDs with those in clinical data
high_sample_ids <- intersect(sample_ids_cna, high_samples$Sample.ID)
low_sample_ids <- intersect(sample_ids_cna, low_samples$Sample.ID)

# Subset data_cna to create high and low groups
high_group <- copy_number_data[, c("Hugo_Symbol", "Entrez_Gene_Id", high_sample_ids)]
low_group <- copy_number_data[, c("Hugo_Symbol", "Entrez_Gene_Id", low_sample_ids)]
```

```{r}
#write.table(high_sample_ids, "/home/evannaughton/project/r_code/high_sample_ids", , sep="\t", quote = F)
#write.table(low_sample_ids, "/home/evannaughton/project/r_code/low_sample_ids", , sep="\t", quote = F)
```

```{r}
# Perform statistical test for each gene
p_values <- numeric(nrow(high_group))
for (i in 1:nrow(high_group)) {
  high_values <- high_group[i, -c(1, 2)]
  low_values <- low_group[i, -c(1, 2)]
  p_values[i] <- t.test(high_values, low_values)$p.value
}

# Adjust p-values for multiple testing
p_values_adjusted <- p.adjust(p_values, method = "BH")

summary(p_values_adjusted)
```

The minumum p-value is 0.0692, indicating no statistical significance between any of the genes between the high and low groups in terms of copy number. Could increase the p-value, but this increases the chance of false positives.
 
```{r}
# Identify most significant genes
significant_genes <- high_group[p_values_adjusted < 0.1, ]
```

```{r}
# Seeing if any genes in the up_list or down_list are present in the most significantly DE copy number genes

# Check if genes in up_list are present in Hugo_Symbol column
up_genes_present <- significant_genes$Hugo_Symbol %in% up_list

# Check if genes in down_list are present in Hugo_Symbol column
down_genes_present <- significant_genes$Hugo_Symbol %in% down_list

# Print genes present in up_list
up_genes_in_significant <- significant_genes$Hugo_Symbol[up_genes_present]
print(up_genes_in_significant)

# Print genes present in down_list
down_genes_in_significant <- significant_genes$Hugo_Symbol[down_genes_present]
print(down_genes_in_significant)
```

When the p-value is set to 1, the only gene upregulated in the high samples that has a 'significant' difference between high and low groups in terms of copy number is MYH4.

CNA scores betwen quartiles:

```{r}
# Read the CNA data 
cna_data <- read.delim("/home/evannaughton/project/acc_tcga/data_cna.txt", check.names = FALSE)
cna_hg19 <- read.delim("/home/evannaughton/project/acc_tcga/data_cna_hg19.seg")

# View the first few rows to confirm the structure
head(cna_data)
head(cna_hg19)
```


```{r}
# Calculating amplification score, deletion score and cna score

# Remove the first two columns (Hugo_Symbol and Entrez_Gene_Id)
cna_data <- cna_data[, -c(1, 2)]

# Convert cna_data to a matrix
cna_matrix <- as.matrix(cna_data)

# Calculate amplification score (number of genes with score 1 or 2)
amplification_score <- colSums(cna_matrix == 1 | cna_matrix == 2)

# Calculate deletion score (number of genes with score -1 or -2)
deletion_score <- colSums(cna_matrix == -1 | cna_matrix == -2)

# Combine scores into a single data frame
scores <- data.frame(
  Sample = colnames(cna_matrix),
  Amplification_Score = amplification_score,
  Deletion_Score = deletion_score
)

# Add the new column "cna_score"
scores <- scores %>%
  mutate(cna_score = Amplification_Score + Deletion_Score)

head(scores)
```

```{r}
# Convert segment means to absolute copy numbers using the correct formula
cna_hg19$abs_copy_number <- 2^(cna_hg19$seg.mean + 1)

# Aggregate by sample to get the mean absolute copy number score
absolute_copy_number_score <- cna_hg19 %>%
  group_by(ID) %>%
  summarise(Mean_Abs_Copy_Number = mean(abs_copy_number, na.rm = TRUE))

# View the absolute copy number scores
head(absolute_copy_number_score)
```

```{r}
# Merge with the previous scores
final_scores <- scores %>%
  left_join(absolute_copy_number_score, by = c("Sample" = "ID"))

# View the final scores
head(final_scores)
```

```{r}
# Rename the Sample column to Sample.ID
colnames(final_scores)[colnames(final_scores) == "Sample"] <- "Sample.ID"

# Merging absolute CNA scores to our finaldf dataframe

# Merged dataframes based on Sample_ID
finaldf <- merge(finaldf, final_scores, by = "Sample.ID", all = TRUE)

# Remove rows with NA only in the Study.ID column
finaldf <- subset(finaldf, !is.na(Study.ID))

# Reset row indices
rownames(finaldf) <- NULL
```

```{r}
# Separating the data based on quartiles

quartile_data <- finaldf %>% 
  filter(quartile %in% c(1, 4))

# Turning quartile into a factor

quartile_data$quartile <- as.factor(quartile_data$quartile)
```

```{r}
# boxplot (cna score)

pCNA <- ggplot(quartile_data, aes(x = quartile, y = cna_score, color = quartile)) +
  geom_boxplot(alpha = 0) +  
  geom_point(position = position_jitter(width = 0.2), size = 3) +  
  scale_color_manual(values = c("turquoise", "red")) +
  labs(x = "Quartile (1 = low MKI67, 4 = high MKI67)", y = "CNA Score") +
  ggtitle("CNA Score per MKI67 Expression Quartile") +
  theme_minimal()

# Calculate p-value between Quartile 1 and Quartile 4
p_value <- t.test(cna_score ~ quartile, data = quartile_data)$p.value

# Add p-value to the plot
pCNA <- pCNA + annotate("text", x = 1.5, y = max(quartile_data$cna_score), 
                  label = paste("p-value =", round(p_value, 4)), 
                  hjust = 0.5, vjust = 1, size = 5, color = "black")

print(pCNA)

# Perform t-test
t_test_result <- t.test(cna_score ~ quartile, data = quartile_data)

print(t_test_result)
```

```{r}
# boxplot (mean absolute cna score)
p2 <- ggplot(quartile_data, aes(x = quartile, y = Mean_Abs_Copy_Number, color = quartile)) +
  geom_boxplot(alpha = 0) +  
  geom_point(position = position_jitter(width = 0.2), size = 3) +  
  scale_color_manual(values = c("turquoise", "red")) +
  labs(x = "Quartile (1 = low MKI67, 4 = high MKI67)", y = "Mean Absolute CNA Score") +
  ggtitle("Mean Absolute CNA Score per MKI67 Expression Quartile") +
  theme_minimal()

# Calculate p-value between Quartile 1 and Quartile 4
p_value <- t.test(Mean_Abs_Copy_Number ~ quartile, data = quartile_data)$p.value

# Add p-value to the plot
p2 <- p2 + annotate("text", x = 1.5, y = max(quartile_data$Mean_Abs_Copy_Number), 
                  label = paste("p-value =", round(p_value, 4)), 
                  hjust = 0.5, vjust = 1, size = 5, color = "black")

print(p2)

# Perform t-test
t_test_result <- t.test(Mean_Abs_Copy_Number ~ quartile, data = quartile_data)

print(t_test_result)
```

```{r}
# boxplot (amplification score)
p3 <- ggplot(quartile_data, aes(x = quartile, y = Amplification_Score, color = quartile)) +
  geom_boxplot(alpha = 0) +  
  geom_point(position = position_jitter(width = 0.2), size = 3) +  
  scale_color_manual(values = c("turquoise", "red")) +
  labs(x = "Quartile (1 = low MKI67, 4 = high MKI67)", y = "Amplification Score") +
  ggtitle("Amplification Score per MKI67 Expression Quartile") +
  theme_minimal()

# Calculate p-value between Quartile 1 and Quartile 4
p_value <- t.test(Amplification_Score ~ quartile, data = quartile_data)$p.value

# Add p-value to the plot
p3 <- p3 + annotate("text", x = 1.5, y = max(quartile_data$Amplification_Score), 
                  label = paste("p-value =", round(p_value, 4)), 
                  hjust = 0.5, vjust = 1, size = 5, color = "black")

print(p3)

# Perform t-test
t_test_result <- t.test(Amplification_Score ~ quartile, data = quartile_data)

print(t_test_result)
```

```{r}
# boxplot (deletion score)
p4 <- ggplot(quartile_data, aes(x = quartile, y = Deletion_Score, color = quartile)) +
  geom_boxplot(alpha = 0) +  
  geom_point(position = position_jitter(width = 0.2), size = 3) +  
  scale_color_manual(values = c("turquoise", "red")) +
  labs(x = "Quartile (1 = low MKI67, 4 = high MKI67)", y = "Deletion Score") +
  ggtitle("Deletion Score per MKI67 Expression Quartile") +
  theme_minimal()

# Calculate p-value between Quartile 1 and Quartile 4
p_value <- t.test(Deletion_Score ~ quartile, data = quartile_data)$p.value

# Add p-value to the plot
p4 <- p4 + annotate("text", x = 1.5, y = max(quartile_data$Deletion_Score), 
                  label = paste("p-value =", round(p_value, 4)), 
                  hjust = 0.5, vjust = 1, size = 5, color = "black")

print(p4)

# Perform t-test
t_test_result <- t.test(Deletion_Score ~ quartile, data = quartile_data)

print(t_test_result)
```

